The Talking Heads Experiment

### 2015-04-16 chap2/chap2.tex
hyphenation, orphans, widows and some other layout updated

However, in the case of computer-based \is{teleportation}
artificial agents, we {\itshape can} 
make the distinction. It is possible to capture the mental state
of an agent in software, load this into a physical 
body, and then operate 
the agent. Afterwards the agent can extract himself
again from the body, teleport himself 
to another physical location through a data transmission network
like the Internet, get instantiated there in another
body, and experience another reality and physically meet
other agents. This is exactly how we have implemented the 
Talking Heads experiment.\footnote{
The agent teleportation infrastructure is in itself a 
fascinating non-trivial engineering project. Contributions 
from Angus McIntyre, Alexis Agahi, Sylvere Tajan and
Frederic Kaplan are gratefully acknowledged (\citealt{McIntyre:1999}). 
The fact that agents can teleport proves that we are dealing 
with a truly distributed multi-agent system. It also 
introduces physical parallelism in the agent-agent and 
agent-environment interactions. 
For a general introduction into multi-agent system
technologies and design methodologies, 
see \cite{Ferber:1999}.}
There are on the one hand
the physical structures, which I will refer to as the {\it
robot bodies}. They are installed in different physical 
locations somewhere in the world and connected
with each other through the Internet. Then there is 
a population of software structures that 
are occasionally loaded and instantiated in specific
robots. I will call these software structures
{\itshape virtual agents}. 
A {\itshape real agent} (a Talking Head) 
only exists when the virtual agent is
loaded in a physical robot body.

@@

The teleporting setup enables some fascinating 
experiments, engulfing the 
whole globe. The same agent can look at a scene from
different points of view or at scenes in different
physical locations in the world by teleporting himself in different 
bodies. He can develop categories in one location
and enrich his learning experience by moving to a new 
location which has different objects and thus poses new 
categorial challenges.
There can be populations of varying sizes with new 
agents being born and older agents dying, just like in 
natural human populations, so that we can study the 
transmission of language from one generation to the next
or the resilience of a language against an influx and 
outflux of agents. We can also let agents develop in
different parts of the world and have them migrate to
study intercultural exchange and language contact.

@@

Concretely, each robot consists of five building blocks (see \figref{f:plate2}):
\begin{itemize}
\item A camera mounted on pan/tilt motors so that it can move 
up or down and left or right. 
\item A loudspeaker (for voice output) and a microphone (for 
voice input). Each agent has a particular quality of voice
output with male and female voices, so that it is possible 
to keep them apart in the dialogue. 
\item A computer that can house an agent's 
cognitive architecture as well as peripheral 
control software to steer the movements of the camera, 
receive and preprocess images, or synthesise and analyse 
sound. This computer is connected 
to the Internet to allow virtual agents to be loaded and instantiated. 
\item A television screen that shows {\itshape us} what the agent
instantiated in a body currently sees through the camera. 
\item A computer screen that shows {\itshape us} what is going on inside
the brain of the agent currently installed in the robot (\figref{f:agentview})
\end{itemize}

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.60\textwidth]{chap1/figs/Interface.pdf}}
\caption{Interface through which the internal states of agents can 
be inspected. Two agents are shown. The top windows show the state of an agent, 
the middle windows the camera inputs that the agent sees and the bottom windows show 
their discrimination trees.}
\label{f:agentview}
\end{figure}

@@

\section{The Guessing Game}

@@

The speaker first looks at one area of the white board and 
directs the attention of the hearer to the same area.\footnote{
I will explain later how exactly agents
decide on a particular scene and how they are able to 
draw each other's attention to specific areas of the 
white board.}
The objects located in this area constitute the {\itshape context}. 
The speaker then chooses one object from the context, which 
I will call the {\itshape topic}, and gives a verbal hint
to the hearer. The {\itshape verbal hint} is an expression
that identifies the topic 
with respect to the other objects in the context. For example, 
if the context contains a red square, a blue triangle, 
and a green circle, then the speaker may say something like 
`the red one' to identify the red square as the topic. 
If the context contains also a red triangle, he has to be more 
precise and say something like `the red square' to delineate
the topic from the red triangle as well as
from the blue square. 
Of course, the Talking Heads do not say `the red square'
but use their own language and concepts which are not 
necessarily the same as those used in English. For example, 
they may say `malewina' to mean [UPPER EXTREME-LEFT LOW-REDNESS].

@@

The environment of the Talking Heads is 
not fixed. The human experimenter 
may change the position of objects, add new kinds of 
objects, or eliminate
others. Consequently a strategy of naming individual objects
will not work. It would lead to a proliferation of 
proper names and it would require the Talking Heads to
recognise objects, which is very difficult to do.\footnote{
For a thorough exposition of the difficulties of 
object recognition, see \cite{Ullman:1996}. 
Object constancy comes fairly late in the 
acquisition of a child's ontology, as Piaget's conservation
experiments have shown. Language probably plays an important
role in forming the notion of an object.}
Indeed, humans don't exclusively use proper names in 
natural language conversations 
either. We say `could you give me the red small square' as 
opposed to `could you give me O\_143'. Natural language 
words like `red' or `small' name perceptually grounded categories
and syntactic structures indicate how they should
be combined and used to find the topic. The relation between 
a language expression and its referent is therefore 
always indirect. This is summarised
in the semiotic square (\figref{triangle}), which I will use throughout this book to 
help understand and analyse the nature of language communication. \is{semiotic square}
The semiotic square relates the four entities
involved in a verbal interaction: 
\begin{itemize}
\item An {\itshape utterance}, such as `small red square', which is 
transmitted as a physical 
signal from one agent to another one through the 
external environment. It is written between double quotes. 
\item A {\itshape meaning}, which consists of categories like [RED], [SMALL], 
or combinations of categories, like \{[RED] [SMALL]\}. Labels of 
categories are written in capital letters between 
square brackets. 
\item An {\itshape image segment}, denoted as $S143$, which is 
a segment of the image perceived through the camera.
\item A {\itshape referent}, like object $O143$, which is 
an entity in the real world. 
\end{itemize}

@@

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.85\textwidth]{chap2/figs/triangle2.pdf}}
\caption{\label{triangle2a}Left: processes carried
out by the speaker. Right: processes carried out by the hearer.
There are also feedback processes moving in alternate directions
until the agents settle on coherent choices for all the items
in their semiotic squares.}
\end{figure}

\item The speaker must then {\itshape conceptualise} the 
scene on the basis of this perception. He must find a set of
categories or a conjunctive combination of 
categories that distinguishes the referent from the other objects in 
the context, and which will act as the meaning of 
his communication. For example, he might choose [BLUE] if the
topic has a blue colour and all the other objects in the 
context are not blue.

\item The speaker must then {\itshape verbalise} this conceptualisation. 
He must use his language system to find words and 
syntactic constructions expressing this meaning.
For example, he might choose `blauw' (if he speaks
Dutch), or `bleu' (if he speaks French),
to convey the category [BLUE].

@@

\begin{table}[htbp]
\caption{}
\label{tab:2:findthereferent}
\begin{tabular}{ l  l  l  l }
\lsptoprule
{\itshape Activity} & {\itshape From} & {\itshape To} & {\itshape Knowledge source}\\ \midrule
Perceive & Object & Image segments & Visual processes \\ 
Conceptualise & Topic & Meaning & Ontology  \\ 
Verbalise & Meaning & Utterance & Lexicon + syntax \\  
Interpret & Utterance & Meaning & Lexicon + syntax \\ 
Apply & Meaning & Topic & Ontology \\ 
Act & Topic & Object & Behavioural process  \\ 
\lspbottomrule
\end{tabular}
\end{table}
\todo[inline]{This table needs a caption}
There is no simple linear flow from perception to utterance 
or from utterance to perception. Instead, we must imagine a dynamic process 
involving forward and backward propagation of information 
until coherent choices for all the nodes in the two
semiotic squares have been established by the speaker and 
the hearer. Many different choices are initially possible 
(many segmentations, conceptualisations, verbalisations, 
interpretations), but the dynamic process gradually settles into 
a single coherent attractor, so that speaker and hearer 
agree upon a common referent. The co-ordination between 
chosen topic and identified referent is done through 
the real world (pointing, handing an object, performing
an action).

@@

Next low-level visual processes gather information about each \is{sensory channels}
segment, such as its average colour, size, shape,
the position with respect to horizontal or vertical axes. 
Each process outputs its information on a {\itshape sensory 
channel} scaled between 0.0 and 1.0. The mechanisms from 
the conceptualisation layer that subsequently use this information
can operate on any kind of sensory channel. 
I will assume for the rest of this chapter that the low-level
routines produce each only three types of information, sent
on the following sensory channels: 
\begin{itemize}
\item The first channel is called HPOS (horizontal position). 
It contains the x-midposition of a segmented object within
the field of view of the robot. 
\item The second channel is called VPOS (vertical position). It
specifies the y-midposition of the segmented object. 
\item The third channel is called GRAY and contains 
the average gray-scale of the object. 
\end{itemize}
Later on additional channels will be introduced.

Consider the two objects in \figref{scene1-1}. The triangular
object has the (scaled) values HPOS=0.35, VPOS=0.40, GRAY=0.33, and 
the rectangular object has the values HPOS=0.70, VPOS=0.85, 
GRAY=0.33. The agents can already visually distinguish
millions of possible scenes with these three sensory
channels. The number of scenes
quickly grows when the set of sensory channels increases.

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.50\textwidth]{chap2/figs/scene1-1.pdf}}
\caption{\label{scene1-1}The scene contains two 
objects: a triangular object and a rectangular object
with the same gray levels. 
Each one is characterised by values on three sensory 
channels: HPOS, VPOS, and GRAY.}
\end{figure}

@@

The data on sensory channels are values from a continuous domain
(between 0.0 and 1.0). 
To be the basis of natural language communication, these
values must be transformed into a discrete domain. This is 
precisely the task of the conceptual layer. It 
performs massive data reduction to make infinitely
rich environments managable. One means of categorisation
is to divide up each domain of values on a particular
sensory channel into regions and 
simply assign a category to each region, thus creating a 
discrimination tree.\is{discrimination tree} For example, 
the HPOS-channel can be cut in two halves leading to a 
distinction between [LEFT] ($0.0 \leq HPOS < 0.5$) and [RIGHT] 
($0.5 \leq HPOS \leq 1.0$). The triangular object 
in \figref{scene1-1} has the value HPOS=0.35 
and would therefore be categorised as [LEFT]. Similarly, the 
VPOS-channel can be divided in two halves yielding the 
categories [LOWER] and [UPPER], and likewise the GRAY-channel
yielding the categories [LIGHT] and [DARK]. 
Given these categories, 
the rectangular object in the scene of 
\figref{scene1-1} would be categorised as [RIGHT UPPER LIGHT].
Of course, light, dark, left, lower, etc. are labels 
that I have given to these categories. The Talking Heads 
create categories by partitioning sensory channels but do
not use these labels internally.

It is always possible to refine a distinction by further
subdividing its region. Thus an agent could further divide the 
bottom region of the HPOS-channel (categorised as 
[LEFT]) into two subregions [TOTALLY-LEFT] ($0.0 \le HPOS < 0.25$), 
and [MID-LEFT] ($0.25 \le HPOS < 0.5$). The triangular object
can now be categorised as [MID-LEFT], if simply categorising it
as [LEFT] is not distinctive enough. Each of these categories can 
still be further refined. The categorisation networks
resulting from these consecutive binary subdivisions form
discrimination trees (\figref{tree1}). It is not at
all assumed that all agents have the same trees; due to 
different developmental histories, divergence is inevitable.

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.35\textwidth]{chap2/figs/tree1.pdf}}
\caption{\label{tree1}A discrimination tree displays
the divisions of the total range of 
values on a sensory channel into finer and finer subregions. 
Categories are assigned to each region at different 
levels of the tree.}
\end{figure}

@@

We will see that agents build hundreds,
even thousands, of categories as they play their
language games, and in addition they make combinations of 
categories. To bring some order in this profusion of 
categories, I will label them using the 
sensory channel from which a category operates, 
followed by the upper and lower bound of the region they 
carve out. Thus [TOTALLY-LEFT]
is labeled as [HPOS 0.0–0.25], because it carves 
out a region between 0.0 and 0.25 on the HPOS-channel. 
When it must be emphasised that 
a category belongs to a particular agent, for example {\bfshape a1}, we
write [HPOS 0.0–0.25]$_{\mathbf a1}$. The same category in agent {\bfshape a2}
is labeled as [HPOS 0.0–0.25]$_{\mathbf a2}$. If I want to talk about 
this category in the abstract, I will not mention any 
agent and simply write [HPOS 0.0–0.25]. We will also 
allow conjunctive combinations of categories which will be
written as a set, as in \{[HPOS 0.0–0.25], [VPOS 0.0–0.25]\}, 
which could be paraphrased as totally-left {\itshape and}
totally-down.

@@

Here are a few example games to give a general 
flavor of this selectionist approach to word acquisition. 
Let us assume that there are only two 
agents, {\bfshape a1} and {\bfshape a2}, and they use only the 
three sensory channels introduced earlier: VPOS for 
vertical position, HPOS for horizontal position, and GRAY
for grayscale.

@@

Here is the simplest possible instance of a
language game, based on the scene in \figref{scene1-1}. 
The speaker, {\bfshape a1}, has picked the triangle as the topic.
To give a verbal hint, he needs 
to conceptualise this topic, which means in this 
specific case, to find a 
category (or set of categories) which distinguishes the 
topic from the other objects in the context. Here 
the context contains only one additional 
object, namely the rectangle. The category
[VPOS 0.0–0.5]$_{\mathbf a1}$ (lower) fits the criteria. 
[VPOS 0.0–0.5] is valid when VPOS$ < 0.5$
which is the case for the triangle but not for the rectangle. 
{\bfshape a1} has an association in his lexicon 
relating [VPOS 0.0–0.5]$_{\mathbf  a1}$ with the word `lu', 
{\bfshape a1} retrieves this word and transmits it to 
the hearer, which is agent {\bfshape a2}.

{\bfshape a2} has stored in his lexicon an
association between `lu' and [VPOS 0.0–0.5]$_{\mathbf  a2}$, and so
hypothesises that [VPOS 0.0–0.5]$_{\mathbf  a2}$ must be the meaning of `lu'. 
When this category is applied to the present scene, in other words,
when {\bfshape a2} filters out the objects whose value for VPOS do not 
fall in the region $[0.0–0.5]$, only one 
remaining object is obtained, the triangle. Hence {\bfshape a2} concludes that 
this must be the topic and points to it.

@@

There are (at least) two ways in which a game similar to game 125, 
but played earlier might have failed. First of all it can be that
the speaker does not have a word yet for the meaning he wants
to convey. The speaker then invokes a strategy 
to repair this failure. The simplest strategy is to 
create a new word for the present meaning. This is how 
{\bfshape a1} might have created the word `lu', and associated it 
with [VPOS 0.0–0.5] in his lexicon. Such simple constructive steps cause
new words to enter the lexicon.

Second, it can be that the hearer does not know the word. The 
game then fails and the speaker points to 
the topic so that the hearer can make 
an educated guess what the meaning might have been:
If the hearer is able to recover a possible topic
from the non-verbal hint given by the speaker, 
he himself can seek a distinctive category or set of categories 
that delineates this topic from the other objects in 
the context just as the speaker has done. It is 
possible (although not necessary as we will see)
that the hearer {\bfshape a2} arrives at his 
own version of the same category, namely [VPOS 0.0–0.5]$_{\mathbf  a2}$. The 
hearer now stores in his lexicon a new association between the 
form heard, `lu', and the guessed meaning [VPOS 0.0–0.5]$_{\mathbf  a2}$. 
Based on this extended lexicon, he will succeed in the same game 
in the future and he can use `lu' himself to verbalise [VPOS 0.0–0.5]$_{\mathbf  a2}$
when talking to other agents. This is the way new words
spread in the population.

@@

The observant reader will have noticed immediately that 
I have swept some important difficulties under the carpet. 
First, it could very well be that unknown to the speaker 
another agent had already used the word `lu' 
for another meaning, so that there are now 
two alternative meanings for `lu' in the lexicon. `lu' 
has become ambiguous. Second, it may be the hearer {\bfshape a2} already 
had a word for [VPOS 0.0–0.5]$_{\mathbf  a2}$, for example `bomida', 
so that there are now two synonyms for [VPOS 0.0–0.5] in the lexicon. 
Synonymy and ambiguity are very common in 
natural languages and unavoidable when a group of 
distributed autonomous agents creates a language without
a central co-ordinator. This implies that 
the agents' lexicon must be sophisticated enough 
to support multiple associations. An agent must be able to 
store the different meanings being used for the same word, and the 
different words being used for the same meaning.

@@

Another difficulty which I did not deal with when discussing Game 25 above,
is that the hearer may conceptualise
the scene differently from the speaker. For example, 
the triangular object is not only located at the lower half
of the scene and the rectangle at the upper half, 
but it is also to the left, i.e. with $HPOS < 0.5$, whereas
the rectangle is to the right. It follows that the 
hearer could just as well have hypothesised that `lu' means [HPOS 0.0–0.5] (left)
and not [VPOS 0.0–0.5] (lower).

>>>>

\enlargethispage{1\baselineskip}

However, in the case of computer-based \is{teleportation}
artificial agents, we {\itshape can} 
make the distinction. It is possible to capture the mental state
of an agent in software, load this into a physical 
body, and then operate 
the agent. Afterwards the agent can extract himself
again from the body, teleport himself 
to another physical location through a data transmission network
like the Internet, get instantiated there in another
body, and experience another reality and physically meet
other agents. This is exactly how we have implemented the 
Talking Heads experiment.\footnote{The agent teleportation infrastructure is in itself a 
fascinating non-trivial engineering project. Contributions 
from Angus McIntyre, Alexis Agahi, Sylvere Tajan and
Frederic Kaplan are gratefully acknowledged (\citealt{McIntyre:1999}). 
The fact that agents can teleport proves that we are dealing 
with a truly distributed multi-agent system. It also 
introduces physical parallelism in the agent-agent and 
agent-environment interactions. 
For a general introduction into multi-agent system
technologies and design methodologies, 
see \cite{Ferber:1999}.}
There are on the one hand
the physical structures, which I will refer to as the {\it
robot bodies}. They are installed in different physical 
locations somewhere in the world and connected
with each other through the Internet. Then there is 
a population of software structures that 
are occasionally loaded and instantiated in specific
robots. I will call these software structures
{\itshape virtual agents}. 
A {\itshape real agent} (a Talking Head) 
only exists when the virtual agent is
loaded in a physical robot body.

@@

The teleporting setup enables \todo{'some' was deleted here to fit the line} fascinating 
experiments, engulfing the 
whole globe. The same agent can look at a scene from
different points of view or at scenes in different
physical locations in the world by teleporting himself in different 
bodies. He can develop categories in one location
and enrich his learning experience by moving to a new 
location which has different objects and thus poses new 
categorial challenges.
There can be populations of varying sizes with new 
agents being born and older agents dying, just like in 
natural human populations, so that we can study the 
transmission of language from one generation to the next
or the resilience of a language against an influx and 
outflux of agents. We can also let agents develop in
different parts of the world and have them migrate to
study intercultural exchange and language contact.

@@

Concretely, each robot consists of five building blocks (see \figref{f:plate2}):
\enlargethispage{1\baselineskip}
\begin{itemize}
\item A camera mounted on pan/tilt motors so that it can move 
up or down and left or right. 
\item A loudspeaker (for voice output) and a microphone (for 
voice input). Each agent has a particular quality of voice
output with male and female voices, so that it is possible 
to keep them apart in the dialogue. 
\item A computer that can house an agent's 
cognitive architecture as well as peripheral 
control software to steer the movements of the camera, 
receive and preprocess images, or synthesise and analyse 
sound. This computer is connected 
to the Internet to allow virtual agents to be loaded and instantiated. 
\item A television screen that shows {\itshape us} what the agent
instantiated in a body currently sees through the camera. 
\item A computer screen that shows {\itshape us} what is going on inside
the brain of the agent currently installed in the robot (\figref{f:agentview})
\end{itemize}

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.75\textwidth]{chap1/figs/Interface.pdf}}
\caption{Interface through which the internal states of agents can 
be inspected. Two agents are shown. The top windows show the state of an agent, 
the middle windows the camera inputs that the agent sees and the bottom windows show 
their discrimination trees.}
\label{f:agentview}
\end{figure}

@@

\clearpage
\section{The Guessing Game}

@@

The speaker first looks at one area of the white board and 
directs the attention of the hearer to the same area.\footnote{
I will explain later how exactly agents
decide on a particular scene and how they are able to 
draw each other's attention to specific areas of the 
white board.}
The objects located in this area constitute the {\itshape context}. 
The speaker then chooses one object from the context, which 
I will call the {\itshape topic}, and gives a verbal hint
to the hearer. The {\itshape verbal hint} is an expression
that identifies the topic 
with respect to the other objects in the context. For example, 
if the context contains a red square, a blue triangle, 
and a green circle, then the speaker may say something like 
`the red one' to identify the red square as the topic. 
If the context contains also a red triangle, he has to be more 
precise and say something like `the red square' to delineate
the topic from the red triangle as well as
from the blue square. 
Of course, the Talking Heads do not say `the red square'
but use their own language and concepts which are not 
necessarily the same as those used in English. For example, 
they may say `malewina' to mean [\textsc{upper extreme}-\textsc{left low}-\textsc{redness}].

@@

The environment of the Talking Heads is 
not fixed. The human experimenter 
may change the position of objects, add new kinds of 
objects, or eliminate
others. Consequently a strategy of naming individual objects
will not work. It would lead to a proliferation of 
proper names and it would require the Talking Heads to
recognise objects, which is very difficult to do.\footnote{
For a thorough exposition of the difficulties of 
object recognition, see \cite{Ullman:1996}. 
Object constancy comes fairly late in the 
acquisition of a child's ontology, as Piaget's conservation
experiments have shown. Language probably plays an important
role in forming the notion of an object.}
Indeed, humans don't exclusively use proper names in 
natural language conversations 
either. We say `could you give me the red small square' as 
opposed to `could you give me O\_143'. Natural language \todo{inconsistency: we have 'O\_143' here but 'O143' on the next page}
words like `red' or `small' name perceptually grounded categories
and syntactic structures indicate how they should
be combined and used to find the topic. The relation between 
a language expression and its referent is therefore 
always indirect. This is summarised
in the semiotic square (\figref{triangle}), which I will use throughout this book to 
help understand and analyse the nature of language communication. \is{semiotic square}
The semiotic square relates the four entities
involved in a verbal interaction: 
\begin{itemize}
\item An {\itshape utterance}, such as `small red square', which is 
transmitted as a physical 
signal from one agent to another one through the 
external environment. It is written between double quotes. 
\item A {\itshape meaning}, which consists of categories like [\textsc{red}], [\textsc{small}], 
or combinations of categories, like \{[\textsc{red}] [\textsc{small}]\}. Labels of 
categories are written in capital letters between 
square brackets. 
\item An {\itshape image segment}, denoted as \emph{S143}, which is 
a segment of the image perceived through the camera.
\item A {\itshape referent}, like object \emph{O143}, which is 
an entity in the real world. 
\end{itemize}

@@

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.75\textwidth]{chap2/figs/triangle2.pdf}}
\caption{\label{triangle2a}Left: processes carried
out by the speaker. Right: processes carried out by the hearer.
There are also feedback processes moving in alternate directions
until the agents settle on coherent choices for all the items
in their semiotic squares.}
\end{figure}

\item The speaker must then {\itshape conceptualise} the 
scene on the basis of this perception. He must find a set of
categories or a conjunctive combination of 
categories that distinguishes the referent from the other objects in 
the context, and which will act as the meaning of 
his communication. For example, he might choose [\textsc{blue}] if the
topic has a blue colour and all the other objects in the 
context are not blue.

\item The speaker must then {\itshape verbalise} this conceptualisation. 
He must use his language system to find words and 
syntactic constructions expressing this meaning.
For example, he might choose `blauw' (if he speaks
Dutch), or `bleu' (if he speaks French),
to convey the category [\textsc{blue}].

@@

\begin{table}[htbp]
\caption{}
\label{tab:2:findthereferent}
\begin{tabular}{ l  l  l  l }
\lsptoprule
{\itshape Activity} & {\itshape From} & {\itshape To} & {\itshape Knowledge source}\\ \midrule
Perceive & Object & Image segments & Visual processes \\ 
Conceptualise & Topic & Meaning & Ontology  \\ 
Verbalise & Meaning & Utterance & Lexicon + syntax \\  
Interpret & Utterance & Meaning & Lexicon + syntax \\ 
Apply & Meaning & Topic & Ontology \\ 
Act & Topic & Object & Behavioural process  \\ 
\lspbottomrule
\end{tabular}
\end{table}
\todo{Table 2.1 table needs a caption}
There is no simple linear flow from perception to utterance 
or from utterance to perception. Instead, we must imagine a dynamic process 
involving forward and backward propagation of information 
until coherent choices for all the nodes in the two
semiotic squares have been established by the speaker and 
the hearer. Many different choices are initially possible 
(many segmentations, conceptualisations, verbalisations, 
interpretations), but the dynamic process gradually settles into 
a single coherent attractor, so that speaker and hearer 
agree upon a common referent. The co-ordination between 
chosen topic and identified referent is done through 
the real world (pointing, handing an object, performing
an action).

@@

Next low-level visual processes gather information about each \is{sensory channels}
segment, such as its average colour, size, shape,
the position with respect to horizontal or vertical axes. 
Each process outputs its information on a {\itshape sensory 
channel} scaled between 0.0 and 1.0. The mechanisms from 
the conceptualisation layer that subsequently use this information
can operate on any kind of sensory channel. 
I will assume for the rest of this chapter that the low-level
routines produce each only three types of information, sent
on the following sensory channels: 
\begin{itemize}
\item The first channel is called \textsc{hpos} (horizontal position). 
It contains the x-midpos\-ition of a segmented object within
the field of view of the robot. 
\item The second channel is called \textsc{vpos} (vertical position). It
specifies the y-midpos\-ition of the segmented object. 
\item The third channel is called \textsc{gray} and contains 
the average gray-scale of the object. 
\end{itemize}
Later on additional channels will be introduced.

Consider the two objects in \figref{scene1-1}. The triangular
object has the (scaled) values \textsc{hpos}=0.35, \textsc{vpos}=0.40, \textsc{gray}=0.33, and 
the rectangular object has the values \textsc{hpos}=0.70, \textsc{vpos}=0.85, 
\textsc{gray}=0.33. The agents can already visually distinguish
millions of possible scenes with these three sensory
channels. The number of scenes
quickly grows when the set of sensory channels increases.

\begin{figure}[b]
  \centerline{\includegraphics[width=.50\textwidth]{chap2/figs/scene1-1.pdf}}
\caption{\label{scene1-1}The scene contains two 
objects: a triangular object and a rectangular object
with the same gray levels. 
Each one is characterised by values on three sensory 
channels: \textsc{hpos}, \textsc{vpos}, and \textsc{gray}.}
\end{figure}

@@

The data on sensory channels are values from a continuous domain
(between 0.0 and 1.0). 
To be the basis of natural language communication, these
values must be transformed into a discrete domain. This is 
precisely the task of the conceptual layer. It 
performs massive data reduction to make infinitely
rich environments managable. One means of categorisation
is to divide up each domain of values on a particular
sensory channel into regions and 
simply assign a category to each region, thus creating a 
discrimination tree.\is{discrimination tree} For example, 
the \textsc{hpos}-channel can be cut in two halves leading to a 
distinction between [\textsc{left}] ($0.0 ≤q \textsc{hpos} < 0.5$) and [\textsc{right}] 
($0.5 ≤q \textsc{hpos} ≤q 1.0$). The triangular object 
in \figref{scene1-1} has the value \textsc{hpos}=0.35 
and would therefore be categorised as [\textsc{left}]. Similarly, the 
\textsc{vpos}-channel can be divided in two halves yielding the 
categories [\textsc{lower}] and [\textsc{upper}], and likewise the \textsc{gray}-channel
yielding the categories [\textsc{light}] and [\textsc{dark}]. 
Given these categories, 
the rectangular object in the scene of 
\figref{scene1-1} would be categorised as [\textsc{right upper light}].
Of course, light, dark, left, lower, etc. are labels 
that I have given to these categories. The Talking Heads 
create categories by partitioning sensory channels but do
not use these labels internally.

It is always possible to refine a distinction by further
subdividing its region. Thus an agent could further divide the 
bottom region of the \textsc{hpos}-channel (categorised as 
[\textsc{left}]) into two subregions [\textsc{totally-left}] ($0.0 ≤ \textsc{hpos} < 0.25$), 
and [\textsc{mid-left}] ($0.25 ≤ \textsc{hpos} < 0.5$). The triangular object
can now be categorised as [\textsc{mid-left}], if simply categorising it
as [\textsc{left}] is not distinctive enough. Each of these categories can 
still be further refined. The categorisation networks
resulting from these consecutive binary subdivisions form
discrimination trees (\figref{tree1}). It is not at
all assumed that all agents have the same trees; due to 
different developmental histories, divergence is inevitable. 
\clearpage
\begin{figure}[htbp]
  \centerline{\includegraphics[width=.35\textwidth]{chap2/figs/tree1.pdf}}
\caption{\label{tree1}A discrimination tree displays
the divisions of the total range of 
values on a sensory channel into finer and finer subregions. 
Categories are assigned to each region at different 
levels of the tree.}
\end{figure}

@@

We will see that agents build hundreds,
even thousands, of categories as they play their
language games, and in addition they make combinations of 
categories. To bring some order in this profusion of 
categories, I will label them using the 
sensory channel from which a category operates, 
followed by the upper and lower bound of the region they 
carve out. Thus [\textsc{totally-left}]
is labeled as [\textsc{hpos} 0.0–0.25], because it carves 
out a region between 0.0 and 0.25 on the \textsc{hpos}-channel. 
When it must be emphasised that 
a category belongs to a particular agent, for example {\bfshape a1}, we
write [\textsc{hpos} 0.0–0.25]$_{\mathbf a1}$. The same category in agent {\bfshape a2}
is labeled as [\textsc{hpos} 0.0–0.25]$_{\mathbf a2}$. If I want to talk about 
this category in the abstract, I will not mention any 
agent and simply write [\textsc{hpos} 0.0–0.25]. We will also 
allow conjunctive combinations of categories which will be
written as a set, as in \{[\textsc{hpos} 0.0–0.25], [\textsc{vpos} 0.0–0.25]\}, 
which could be paraphrased as totally-left {\itshape and}
totally-down.

@@

Here are a few example games to give a general 
flavor of this selectionist approach to word acquisition. 
Let us assume that there are only two 
agents, {\bfshape a1} and {\bfshape a2}, and they use only the 
three sensory channels introduced earlier: \textsc{vpos} for 
vertical position, \textsc{hpos} for horizontal position, and \textsc{gray}
for grayscale.

@@

Here is the simplest possible instance of a
language game, based on the scene in \figref{scene1-1}. 
The speaker, {\bfshape a1}, has picked the triangle as the topic.
To give a verbal hint, he needs 
to conceptualise this topic, which means in this 
specific case, to find a 
category (or set of categories) which distinguishes the 
topic from the other objects in the context. Here 
the context contains only one additional 
object, namely the rectangle. The category
[\textsc{vpos} 0.0–0.5]$_{\mathbf a1}$ (lower) fits the criteria. 
[\textsc{vpos} 0.0–0.5] is valid when \textsc{vpos}$ < 0.5$
which is the case for the triangle but not for the rectangle. 
{\bfshape a1} has an association in his lexicon 
relating [\textsc{vpos} 0.0–0.5]$_{\mathbf  a1}$ with the word `lu', 
{\bfshape a1} retrieves this word and transmits it to 
the hearer, which is agent {\bfshape a2}.

{\bfshape a2} has stored in his lexicon an
association between `lu' and [\textsc{vpos} 0.0–0.5]$_{\mathbf  a2}$, and so
hypothesises that [\textsc{vpos} 0.0–0.5]$_{\mathbf  a2}$ must be the meaning of `lu'. 
When this category is applied to the present scene, in other words,
when {\bfshape a2} filters out the objects whose value for \textsc{vpos} do not 
fall in the region $[0.0–0.5]$, only one 
remaining object is obtained, the triangle. Hence {\bfshape a2} concludes that 
this must be the topic and points to it.

@@

There are (at least) two ways in which a game similar to game 125, 
but played earlier might have failed. First of all it can be that
the speaker does not have a word yet for the meaning he wants
to convey. The speaker then invokes a strategy 
to repair this failure. The simplest strategy is to 
create a new word for the present meaning. This is how 
{\bfshape a1} might have created the word `lu', and associated it 
with [\textsc{vpos} 0.0–0.5] in his lexicon. Such simple constructive steps cause
new words to enter the lexicon.

Second, it can be that the hearer does not know the word. The 
game then fails and the speaker points to 
the topic so that the hearer can make 
an educated guess what the meaning might have been:
If the hearer is able to recover a possible topic
from the non-verbal hint given by the speaker, 
he himself can seek a distinctive category or set of categories 
that delineates this topic from the other objects in 
the context just as the speaker has done. It is 
possible (although not necessary as we will see)
that the hearer {\bfshape a2} arrives at his 
own version of the same category, namely [\textsc{vpos} 0.0–0.5]$_{\mathbf  a2}$. The 
hearer now stores in his lexicon a new association between the 
form heard, `lu', and the guessed meaning [\textsc{vpos} 0.0–0.5]$_{\mathbf  a2}$. 
Based on this extended lexicon, he will succeed in the same game 
in the future and he can use `lu' himself to verbalise [\textsc{vpos} 0.0–0.5]$_{\mathbf  a2}$
when talking to other agents. This is the way new words
spread in the population.

@@

The observant reader will have noticed immediately that 
I have swept some important difficulties under the carpet. 
First, it could very well be that unknown to the speaker 
another agent had already used the word `lu' 
for another meaning, so that there are now 
two alternative meanings for `lu' in the lexicon. `lu' 
has become ambiguous. Second, it may be the hearer {\bfshape a2} already 
had a word for [\textsc{vpos} 0.0–0.5]$_{\mathbf  a2}$, for example `bomida', 
so that there are now two synonyms for [\textsc{vpos} 0.0–0.5] in the lexicon. 
Synonymy and ambiguity are very common in 
natural languages and unavoidable when a group of 
distributed autonomous agents creates a language without
a central co-ordinator. This implies that 
the agents' lexicon must be sophisticated enough 
to support multiple associations. An agent must be able to 
store the different meanings being used for the same word, and the 
different words being used for the same meaning.

@@

Another difficulty which I did not deal with when discussing Game 25 above,
is that the hearer may conceptualise
the scene differently from the speaker. For example, 
the triangular object is not only located at the lower half
of the scene and the rectangle at the upper half, 
but it is also to the left, i.e. with $\textsc{hpos} < 0.5$, whereas
the rectangle is to the right. It follows that the 
hearer could just as well have hypothesised that `lu' means [\textsc{hpos} 0.0–0.5] (left)
and not [\textsc{vpos} 0.0–0.5] (lower).

Is this bad? It depends on the situations being encountered
in the future. When the game is played again for the
same scene with {\bfshape a1} saying `lu' to mean [\textsc{vpos} 0.0–0.5], and 
{\bfshape a2} interpreting `lu' as [\textsc{hpos} 0.0–0.5], the game would succeed! 
Communicative success is achieved whenever the 
hearer recovers the referent chosen by the 
speaker; it is not required that they 
use exactly the same meaning. In fact, the hearer can never know
for sure what meaning was initially conceptualised by the speaker and 
neither can the speaker know which meaning was understood 
by the hearer because they cannot look inside each
other's head. The meaning can be quite different, 
as we have just seen, as long as it is compatible.

\clearpage
Even more remarkably, if a topic which is at the lower
portion of the scene, is always 
on the left side and vice-versa, there would always be 
success despite the different meanings of `lu', 
and the players would never discover
that each means something else by `lu'.
